Class-based network segregation, Economic Inequality and Redistributive Preferences across societies

### 2024-05-22 paper1_jiturra.Rmd
save commit

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. For instance, it has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019]. Another group of studies has suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can also imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. It is argued that continuous interaction with diverse people that characterize interpersonal services can reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed can contribute to self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

@@

To summarise, it has been shown that economic inequality increases the dependency of network segregation with class positions, leading the working classes to be equally segregated regardless of income inequality compared to the increasingly diverse networks of the upper classes [@pichler_social_2009;@otero_differences_2023]. Therefore, class-based network segregation in unequal societies might have a weaker association with redistributive preferences. Thus, economic inequality is expected to _mitigate_ the association of class-based network segregation with redistributive preferences ($H_{2}$).

@@

The primary societal characteristic under examination is the current levels of economic inequality, for which we have utilized the Gini Index (post-taxes and transfers) sourced from the World Income Inequality Dataset (WID) [@alvaredo_world_2022]. Additionally, two contextual variables have been incorporated as controls in the multilevel models. Firstly, economic prosperity has been considered to ensure consistency in the estimates for economic inequality, drawing from the World Income Inequality Dataset (WID) [@Schmidt-Catran2016; @finseraas_income_2009]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state has been included [see @edlund_democratic_2015]. This measure, as an index, considers three dimensions: (i) tax revenue as a percentage of gross domestic product (GDP) [@ilo_world_2022]; (ii) welfare generosity as total governmental spending as a share of GDP [@ilo_world_2022]; and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear models account for the hierarchical structure of the data, where individuals are nested within countries. Therefore, our first analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model allows us to assess the intraclass correlation, revealing that 13.7% of the variance in redistributive preferences can be attributed to belonging to higher-level units (countries). Subsequently, the micro-level models are estimated to determine the association between network homogeneity as well as the interaction with social class for testing our _segregation_ hypothesis ($H_1$). Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class to determine whether economic inequality moderates the effect of the micro-level interaction ($H_2$). Therefore, we estimate a three-way cross-level interaction to determine if the current levels of income inequality affect the interactions between network homogeneity and social class.  In the macro-level models, all the individual-level factors are group-mean centered (CWC) to mitigate possible problems of collinearity [@hox_multilevel_2010]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations  [@hox_multilevel_2010].`

@@

```{r fig1, fig.dim=c(10,5),fig.cap='Cross-country comparison of Redistributive preferences and Social Class ',warning=FALSE}
set1 <-  RColorBrewer::brewer.pal(n = 8, name = "Set1")
options(ggplot2.discrete.fill = set1)
options(ggplot2.discrete.colour = set1)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text=ggplot2::element_text(size=15,  family="serif"))
options(scipen=999)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr,ggplot2,haven,ggrepel,gridExtra)  
rm(list=ls())
load(here::here("input/data/proc/study1_country.RData"));df1 <- df2
df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
dfreg <- df1 %>% dplyr::select(
  egal = egal2,
  dclass3res_V,
  dclass3spo_V,
  homclass3_V_res,
  know_total,
  socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  religion,
  partner,
  union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  gdppercapita,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         )

@@

```{r fig3, fig.dim=c(10,5),fig.cap='Bivariate relationships between Income Inequality, Network homogeneity and Redistributive preferences',warning=FALSE}
set1 <-  RColorBrewer::brewer.pal(n = 8, name = "Set1")
options(ggplot2.discrete.fill = set1)
options(ggplot2.discrete.colour = set1)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text=ggplot2::element_text(size=15,  family="serif"))
options(scipen=999)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr,ggplot2,haven,ggrepel,gridExtra)  
rm(list=ls())

@@

dfreg <- df1 %>% dplyr::select(
  egal = egal2,
  dclass3res_V,
  dclass3spo_V,
  homclass3_V_res,
  know_total,
  socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  religion,
  partner,
  union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  gdppercapita,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         # gini_disp=scale(gini_disp),
         )
homo_ineq <- dfreg %>%
  # filter(country3 != "HUN") %>%
  group_by(country2,country3) %>%
  summarise_at(c("homclass3_V_res","egal","gini_disp","d10d1"), mean, na.rm = TRUE) %>% 
  as.data.frame()

@@

texreg(digclas1_VW,stars = c(0.001, 0.01, 0.05, 0.1),single.row = F,
       caption = paste("(\\#tab:table1)","Multilevel models for Class-based network segregation and Redistributive Preferences"),
       custom.note = "Note: Models include sampling weights. Gender, age, and marital status are included as controls. Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,
       custom.coef.map = ccoef,
       symbol = "+",scalebox = 0.60,
      groups = list("Social Class (Ref.= Service Class)" = 3:5, "Household Income (Ref.= Tertile I)"= 6:8,
                    # "Partner's Social Class (Ref.= Service Class)" = 12:14,
      "Homogeneity x Social Class"=10:11),
      custom.gof.rows = list("Controls"=rep("Yes",5)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups","Var: Country (Intercept)","Var: Residual"),
       float.pos = "h!",
      # file = "table1.html"
       use.packages = F)
# performance::check_collinearity(digclas1_VW_6)
# performance::check_model(digclas1_VW_6,check="vif")
```

@@

texreg(digclas1_macro,stars = c(0.001, 0.01, 0.05, 0.1),symbol = "+",scalebox = 0.7,
       caption = paste("(\\#tab:table2)","Multilevel models for Income Inequality, Network segregation and Redistributive Preferences"),
       custom.note = "Note: Models include sampling weights and individual level controls centered within cluster (group mean). Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,single.row = T,
       custom.coef.map = ccoef,
      groups = list("Social Class (Ref.= Service Class)" = 2:3,"Macro-level factors"=4:6, "Homogeneity x Social Class"=7:8, "Homogeneity x Social Class x Income Inequality"=10:11),
       float.pos = "h!",
      custom.gof.rows = list("Controls"=rep("Yes",4)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups",
                           "Var: Country (Intercept)",
                           "Var: Country Homogeneity",
                           "Var: Country Intermediate Class ",
                           "Var: Country Working Class",
                           "Cov: Country (Intercept), Homogeneity",
                           "Cov: Country (Intercept), Intermediate Class ",
                           "Cov: Country (Intercept), Working Class",
                           "Cov: Country Homogeneity, Intermediate Class",
                           "Cov: Country Homogeneity, Working Class",
                           "Cov: Country Intermediate Class, Working Class",
                           "Var: Residual"),
        # file = "table2.html",
       use.packages = F)
```

@@

First, the average scores of the International Socio-Economic Index of Occupational Status (ISEI) [@ganzeboom_three_2003] for each occupation of the position generator are calculated. Second, the ISEI score of the respondent (R's) is subtracted from the average ISEI points of the personal network. For example, if the R's has an ISEI of 80 and the network ISEI is 50, the social distance will be 30 (80 - 50), a "upward" social distance. Another case could be 50 (R's) minus 80 (network), and the average social distance will be - 30 or "downward" social distance. In addition, When the distance is 0, the network is entirely homogeneous.

>>>>

However, class-based explanations of redistributive preferences have mainly been, but not exclusively, focused on the individual or household situation. For instance, it has been claimed that redistributive preferences can be explained through the labor market situation, which comprises access to economic resources and risk exposure [@meltzer_rational_1981;@rehm_risks_2009].
Moreover, as material interests might prevail in scarcity, value-driven motivations are a relevant explanatory factor of redistributive preferences [@Kulin2013;@feldman_humanitarian_2001], becoming salient under greater certainty and weaker under material precariousness [@Maldonadoetal2019].Another group of studies has suggested that given the substantial time workers spend performing their jobs, social relations in the workspace can also imprint normative views that ultimately shape political opinions [@oesch_redrawing_2006]. It is argued that continuous interaction with diverse people that characterize interpersonal services can reinforce egalitarian values [@kitschelt_occupations_2014]. In contrast, the vertical monitoring in managerial occupations and the relevance of autonomy in the case of self-employed can contribute to self-interested and conservative political views [@langsaether_more_2020;@oesch_electoral_2018].

@@

To summarise, it has been shown that economic inequality increases the dependency of network segregation with class positions, leading the working classes to be equally segregated regardless of income inequality compared to the increasingly diverse networks of the upper classes [@pichler_social_2009;@otero_differences_2023]. Consequently, it is expected that  class-based network segregation in unequal societies might have a weaker association with redistributive preferences. Thus, economic inequality is expected to _mitigate_ the association of class-based network segregation with redistributive preferences ($H_{2}$).

<!-- A simplified framework of the hypotheses is shown in Figure X. -->

@@

The primary societal characteristic under examination is the current levels of economic inequality, for which we have utilized the Gini Index (post-taxes and transfers) sourced from the World Income Inequality Dataset (WID) [@alvaredo_world_2022]. Additionally, two contextual variables are incorporated as controls in the multilevel models. Firstly, employing the Gross Domestic Product (GDP) available in the WID ensures consistency in the economic inequality estimates independently of economic property [@Schmidt-Catran2016; @finseraas_income_2009]. Secondly, to account for heterogeneity in institutional arrangements stemming from welfare schemes, a measure of the size of the welfare state is included [see @edlund_democratic_2015]. The measure is a standardized and rescaled indicator from 0 to 100 that combines (i) tax revenue as a percentage of GDP [@ilo_world_2022], (ii) welfare generosity as total governmental spending as a share of GDP [@ilo_world_2022], and (iii) the current level of redistribution [@solt_measuring_2020].

@@

Multilevel linear models account for the hierarchical structure of the data, where individuals are nested within countries. Therefore, the analysis begins with estimating the null model by declaring the nested structure using random intercept. This initial model assesses the intraclass correlation, revealing that 13.5% of the variance in redistributive preferences can be attributed to belonging to higher-level units (countries). Subsequently, the micro-level models are estimated to determine the association between network homogeneity and the interaction with social class to test the _segregation_ hypothesis ($H_1$). Following this, macro-level models are estimated by incorporating random intercepts alongside random slopes for network homogeneity and social class to determine whether economic inequality moderates the effect of the micro-level interaction ($H_2$). Therefore, a three-way cross-level interaction is estimated to determine if income inequality moderates the interaction between network homogeneity and social class [^1]. In the macro-level models, all the individual-level factors are group-mean centered (CWC) to mitigate possible collinearity problems [@hox_multilevel_2010]. Additionally, all the country-level factors have been standardized (z-scores) to ease comparability in the estimations  [@hox_multilevel_2010].

[^1]: Supplementary analyses employing alternative income inequality measures show that the results are robust when using the Inter-decile ratio (D9/D1) and the Top 10/Bottom 50 ratio. Additionally, countries were classified into low, middle-low, middle, middle-high, and high-income inequality groups based on quintiles according to the Gini index. Hence, country-fixed effects regressions are used to control for the cross-country differences and observed and unobserved societal characteristics. The results are consistent with the multilevel estimations.

@@

```{r fig1, fig.dim=c(10,5),fig.cap='Cross-country comparison of Redistributive preferences and Social Class ',warning=FALSE}
set1 <-  RColorBrewer::brewer.pal(n = 8, name = "Set1")
options(ggplot2.discrete.fill = set1)
options(ggplot2.discrete.colour = set1)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text=ggplot2::element_text(size=15,  family="serif"))
options(scipen=999)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr,ggplot2,haven,ggrepel,gridExtra)  
rm(list=ls())
load(here::here("input/data/proc/study1_country.RData"));df1 <- df2
df1$dclass3spo_V <- car::recode(df1$dclass3spo_V,"NA='No information (Missing, No partner)'",levels = c("Service Class (I+II)","Intermediate class (III+IV)","Working Class (V+VI+VII)","No information (Missing, No partner)"))
dfreg <- df1 %>% dplyr::select(
  egal = egal2,
  dclass3res_V,
  # dclass3spo_V,
  homclass3_V_res,
  know_total,
  # socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  # religion,
  partner,
  # union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  gdppercapita,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         )

@@

```{r fig3, fig.dim=c(10,5),fig.cap='Bivariate relationships between Income Inequality, Network homogeneity and Redistributive preferences',warning=FALSE}
set1 <-  RColorBrewer::brewer.pal(n = 8, name = "Set1")
set.seed(1234)
options(ggplot2.discrete.fill = set1)
options(ggplot2.discrete.colour = set1)
ggplot2::theme_set(ggplot2::theme_bw())
ggplot2::theme_update(text=ggplot2::element_text(size=15,  family="serif"))
options(scipen=999)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(dplyr,ggplot2,haven,ggrepel,gridExtra)  
rm(list=ls())

@@

dfreg <- df1 %>% dplyr::select(
  egal = egal2,
  dclass3res_V,
  # dclass3spo_V,
  homclass3_V_res,
  know_total,
  # socialtrust,
  Q03pcm,
  edyears,
  female,
  agenum,
  # religion,
  partner,
  # union,
  workst,
  WEIGHT,
  region,
  "gini_disp"=wid_gini_disp,
  d10d1=wid_p90p10,
  wid_p90p50,
  top10=wid_sharetop10,
  gdppercapita,
  country2, country, country3,
) %>%
  # filter(oecd == "OECD") %>%
  filter(country2 != "SVN") %>% 
  na.omit() %>% 
    mutate(loggdppercapita=scale(gdppercapita/1000),
         edyears2 = edyears ^ 2,
         age2 = agenum ^ 2, 
         # gini_disp=scale(gini_disp),
         )
homo_ineq <- dfreg %>%
  # filter(country3 != "HUN") %>%
  group_by(country2,country3) %>%
  summarise_at(c("homclass3_V_res","egal","gini_disp","d10d1"), mean, na.rm = TRUE) %>% 
  as.data.frame()

@@

texreg(digclas1_VW,stars = c(0.001, 0.01, 0.05, 0.1),single.row = F,
       caption = paste("(\\#tab:table1)","Multilevel models for Class-based network segregation and Redistributive Preferences"),
       custom.note = "\\item Note: Models include sampling weights. Gender, age, and marital status are included as controls. Standard errors in parentheses. %stars",threeparttable = T,
       caption.above = T,leading.zero = T,
       custom.coef.map = ccoef,
       symbol = "+",scalebox = 0.60,
      groups = list("Social Class (Ref.= Service Class)" = 3:5, "Household Income (Ref.= Tertile I)"= 6:8,
                    # "Partner's Social Class (Ref.= Service Class)" = 12:14,
      "Homogeneity x Social Class"=10:11),
      custom.gof.rows = list("Controls"=rep("Yes",5)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups","Var: Country (Intercept)","Var: Residual"),
       float.pos = "h!",
      # file = "table1.html"
       use.packages = F)
# performance::check_collinearity(digclas1_VW_5)
# performance::check_model(digclas1_VW_5,check="vif")
# performance::icc(digclas1_VW_5)
sjPlot::tab_model(digclas1_VW_5, show.se = F, show.std = TRUE, show.stat = F)
```

@@

texreg(digclas1_macro,stars = c(0.001, 0.01, 0.05, 0.1),symbol = "+",scalebox = 0.7,
       caption = paste("(\\#tab:table2)","Multilevel models for Income Inequality, Network segregation and Redistributive Preferences"),
       custom.note = "\\item Note: Models include sampling weights and individual level controls centered within cluster (group mean). Standard errors in parentheses. %stars",
       caption.above = T,leading.zero = T,single.row = F,threeparttable = T,
       custom.coef.map = ccoef,
      groups = list("Social Class (Ref.= Service Class)" = 2:3,"Macro-level factors"=4:6, "Homogeneity x Social Class"=7:8, "Homogeneity x Social Class x Income Inequality"=10:11),
       float.pos = "h!",
      custom.gof.rows = list("Controls"=rep("Yes",4)),include.loglik = FALSE,include.aic = FALSE,
      custom.gof.names = c(NA,NA,"Num. groups",
                           "Var: Country (Intercept)",
                           "Var: Country Homogeneity",
                           "Var: Country Intermediate Class ",
                           "Var: Country Working Class",
                           "Cov: Country (Intercept), Homogeneity",
                           "Cov: Country (Intercept), Intermediate Class ",
                           "Cov: Country (Intercept), Working Class",
                           "Cov: Country Homogeneity, Intermediate Class",
                           "Cov: Country Homogeneity, Working Class",
                           "Cov: Country Intermediate Class, Working Class",
                           "Var: Residual"),
        # file = "table2.html",
       use.packages = F)
```

@@

First, the average scores of the International Socio-Economic Index of Occupational Status (ISEI) [@ganzeboom_new_2010] for each occupation of the position generator are calculated. Second, the ISEI score of the respondent (R's) is subtracted from the average ISEI points of the personal network. For example, if the R's has an ISEI of 80 and the network ISEI is 50, the social distance will be 30 (80 - 50), a "upward" social distance. Another case could be 50 (R's) minus 80 (network), and the average social distance will be - 30 or "downward" social distance. In addition, When the distance is 0, the network is entirely homogeneous.
